---
title: "College Score Admission and Ethnicity"
author: "antelope"
date: "2022-07-31"
output: html_document
---

## Introduction

The SAT test is an effort to increase academic opportunity, however there exist a racial gap in SAT scores and many universities are starting to not require it. In this report, we would like to explore the whether there is a correlation between proportion of ethnicity and average SAT admission score of universities in the United States. Additionally, we would like to know whether there is a difference in the average admission SAT scores of liberal states vs conservative states to better understand significance of SAT and admission between them.

Thesis Statement: 
There is a racial difference in SAT scores between each ethnicity as shown by the correlation between average SAT admission scores and different ethnicity enrollment proportion of university. Additionally, Liberal state universities have a higher admission SAT score than conservative state universities, which may suggest a more emphasis on SAT scores by liberal universities.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, essage=FALSE, warning=FALSE)
library(tidyverse)
library(lubridate)
library(dplyr)
library(tidyr)
library(leaflet) # geoplot
library(htmltools)
library(patchwork)
library(modelr)
```

## Background

The dataset we have selected for this analysis is "U.S Department of Education: College Scoreboard" which includes information on institutional characteristics, enrollment, student aid, costs, and student outcomes during the year 1996-97 through 2020-21.

https://collegescorecard.ed.gov/data/

https://worldpopulationreview.com/state-rankings/most-liberal-states

We will be using the following columns for analysis:

- INSTNM - Institution name

- STABBR - State postcode

- SAT_AVG - Average SAT equivalent score of students admitted

- SATVRMID - Midpoint of SAT scores at the institution (critical reading)

- SATMTMID - Midpoint of SAT scores at the institution (math)

- UGDS - Enrollment of undergraduate certificate/degree-seeking students

- UGDS_WHITE - Total share of enrollment of undergraduate degree-seeking students who are white

- UGDS_BLACK - Total share of enrollment of undergraduate degree-seeking students who are black

- UGDS_HISP - Total share of enrollment of undergraduate degree-seeking students who are Hispanic

- UGDS_ASIAN - Total share of enrollment of undergraduate degree-seeking students who are Asian

- UGDS_AIAN - Total share of enrollment of undergraduate degree-seeking students who are American Indian/Alaska Native

- UGDS_NHPI - Total share of enrollment of undergraduate degree-seeking students who are Native Hawaiian/Pacific Islander

- UGDS_2MOR - Total share of enrollment of undergraduate degree-seeking students who are two or more races

- UGDS_NRA - Total share of enrollment of undergraduate degree-seeking students who are non-resident aliens


Unusual factors:

- All of the score and ethnicity proportion columns are type char which will need to be converted to numeric for analysis

The following report includes graphical information about the basic information such as distribution of SAT and ACT scores, and distribution of ethnicity. it includes a bar chart showing which schools have highest Average SAT and ACT scores. Then, we will be focusing on whether liberal states or conservative states perform better and the degree of correlation between average score and ethnicity.  

```{r read data, include=FALSE}
# read data
y2017 = read_csv("MERGED2017_18_PP.csv")
y2018 = read_csv("MERGED2018_19_PP.csv")
y2019 = read_csv("MERGED2019_20_PP.csv")
y2020 = read_csv("MERGED2020_21_PP.csv")
```

```{r merge data, include=FALSE}
# merge data
file = y2017 %>%
  merge(y2018, all=TRUE) %>%
  merge(y2019, all=TRUE) %>%
  merge(y2020, all=TRUE)
```

Here we focus on the SATs, ACTs, and ethnicity of a student

```{r, include=FALSE}
# data selection
college = file %>% 
  select("INSTNM", "STABBR", "SAT_AVG", "SATVRMID", "SATMTMID", "ACTENMID", "ACTMTMID", "UGDS_WHITE", "UGDS_BLACK", "UGDS_HISP", "UGDS_ASIAN", "UGDS_AIAN", "UGDS_NHPI", "UGDS_2MOR", "UGDS_NRA", "UGDS")
```

Some university do not have the SATs, ACTs score, and information about ethnicity of students, so we are going to drop those schools out

```{r, include=FALSE}
college[college == 'NULL'] = NA
colSums(is.na(college))
college = college %>% na.omit()

# convert data type to numeric
college = college %>%
  mutate_at(vars("SAT_AVG", "SATVRMID", "SATMTMID", "UGDS_WHITE", "UGDS_BLACK", "UGDS_HISP", "UGDS_ASIAN", "UGDS_AIAN", "UGDS_NHPI", "UGDS_2MOR", "UGDS_NRA", "UGDS"), as.numeric)

# mutate to get population for each race
college = college %>% 
  mutate(white = round(UGDS_WHITE * UGDS),
         black = round(UGDS_BLACK * UGDS),
         asian = round(UGDS_ASIAN * UGDS),
         american_indian = round(UGDS_AIAN * UGDS),
         pacific_islander = round(UGDS_NHPI * UGDS),
         two_or_more = round(UGDS_2MOR * UGDS),
         non_resident_alien = round(UGDS_NRA * UGDS))
```

```{r, include=FALSE}
# data preparation for geo plot
viz = file %>%
  select("INSTNM", "STABBR", "SATVRMID", "SATMTMID", "UGDS_WHITE", "UGDS_BLACK", "UGDS_HISP", "UGDS_ASIAN", "UGDS_AIAN", "UGDS_NHPI","UGDS_2MOR","UGDS_NRA", "ZIP", "LATITUDE", "LONGITUDE", "CITY")

viz[viz == 'NULL'] = NA

uniInfo = paste(viz[['INSTNM']], "<br>", 
                viz[['CITY']], ", ", 
                viz[['STABBR']], viz[['ZIP']], "<br>",
                
                "Avg SAT Math", viz[['SATMTMID']], "<br>", 
                "Avg SAT Verbal", viz[['SATVRMID']], "<br>",
                "White Proportion", as.numeric(viz[['UGDS_WHITE']]) * 100, "%<br>",
                "Black Proportion", as.numeric(viz[['UGDS_BLACK']]) * 100,"%<br>",
                "Asian Proportion", as.numeric(viz[['UGDS_ASIAN']]) * 100,"%<br>",
                "American Indian/Alaska Native Proportion", as.numeric(viz[['UGDS_AIAN']]) * 100,"%<br>",
                "Native Hawaiian/Pacific Islander", as.numeric(viz[['UGDS_NHPI']]) * 100,"%<br>",
                "two or more races", as.numeric(viz[['UGDS_2MOR']]) * 100, "%<br>",
                "non-resident aliens", as.numeric(viz[['UGDS_NRA']]) * 100, "%<br>")

viz$info = uniInfo

## filter data
viz = filter(viz,
              is.na(LATITUDE)==FALSE &
              is.na(LONGITUDE)==FALSE &
              is.na(SATVRMID) == FALSE &
              is.na(SATMTMID) == FALSE)

blingIcon = makeIcon(
  iconUrl = "https://img.icons8.com/color/344/school-building.png",
  iconWidth = 30, iconHeight = 30,
)
```


## Analysis
This is the map showing the proportion of each ethnicity and SAT/ACT scores according to each institution

```{r, echo=FALSE}
# visualize map of university 
map = leaflet(viz) %>% 
          setView(-93.65, 42.0285, zoom = 4) %>%
          addTiles() %>%
          addMarkers(~as.numeric(LONGITUDE), ~as.numeric(LATITUDE), popup=~info,
                     options = popupOptions(closeButton = TRUE),
                     clusterOptions = markerClusterOptions(), 
                     icon = blingIcon)
map
```

```{r, echo=FALSE}
pop_dis = college %>%
  select("INSTNM", "white", "black", "asian", "american_indian", "pacific_islander", "two_or_more", "non_resident_alien") %>%
  group_by(INSTNM) %>%
  summarize(across(everything(), sum)) %>%
  pivot_longer(!INSTNM, names_to = "race", values_to = "population")

ggplot(pop_dis, aes(x=race, y=log10(population), color=race, fill=race, group=race)) +
  geom_violin(scale = "area") +
  ggtitle("Ethnicity Population Distribution across all Institution") +
  xlab("Race") +
  ylab("Population (log scaled 10^7)") + 
  coord_flip()
```


Population Distribution for each races

- ethnicity with highest population is white


```{r average SATs score for each subjects, echo=FALSE}
sat = college %>%
  select("INSTNM", "SATVRMID", "SATMTMID") %>%
  pivot_longer(!INSTNM, names_to = "Section", values_to = "Score")

ggplot(sat, aes(x=Score, color=Section, fill=Section, group=Section)) +
  geom_density(alpha=0.3) +
  ggtitle("Density of Average SAT score across all Institution") +
  xlab("SAT Score") +
  ylab("")
```

Average SAT score :

- The distribution of SAT math and SAT verbal is quite similar

- Achieving perfect SAT verbal is more rare


```{r school with top SAT Math score, echo=FALSE}
top_math = college %>%
  group_by(INSTNM) %>%
  summarize(mean_math = round(mean(SATMTMID))) %>%
  arrange(desc(mean_math)) %>%
  slice_max(mean_math, n = 20)

ggplot(top_math, aes(x=reorder(INSTNM, mean_math), y=mean_math)) +
  geom_bar(stat="identity", fill="lightblue") +
  geom_text(aes(x=INSTNM, y=mean_math-15, hjust=1, label=mean_math), size=3.5) + 
  coord_flip() + 
  ggtitle("Institution with highest average SAT Math score") +
  xlab("Average SAT Math Score") +
  ylab("Institution")
```

```{r school with top SAT Verbal, echo=FALSE}
top_verbal = college %>%
  group_by(INSTNM) %>%
  summarize(mean_verbal = round(mean(SATVRMID))) %>%
  arrange(desc(mean_verbal)) %>%
  slice_max(mean_verbal, n = 20)

ggplot(top_verbal, aes(x=reorder(INSTNM, mean_verbal), y=mean_verbal)) +
  geom_bar(stat="identity", fill="lightgreen") +
  geom_text(aes(x=INSTNM, y=mean_verbal-15, hjust=1, label=mean_verbal), size=3.5) + 
  coord_flip() +
  ggtitle("Institution with highest average SAT Verbal score") +
  xlab("Institution") +
  ylab("Average SAT Verbal Score")
```


```{r proportion of ethnicity in school with lowest admission rate, echo=FALSE}

# proportion of ethnicity in school with lowest admission rate
prop_ad = file %>%
  select("INSTNM", "ADM_RATE", "UGDS_WHITE", "UGDS_BLACK", "UGDS_HISP", "UGDS_ASIAN","UGDS_NHPI","UGDS_2MOR","UGDS_NRA", "UGDS_AIAN") 

prop_ad[prop_ad == 'NULL'] = NA
prop_ad = prop_ad %>% na.omit()

prop_ad = prop_ad %>%
  group_by(INSTNM) %>% 
  slice_min(ADM_RATE) %>%
  filter(ADM_RATE != 0 & INSTNM != "Chicago School of Professional Psychology-College of Nursing") %>%
  arrange(ADM_RATE) %>%
  pivot_longer(!INSTNM & !ADM_RATE, names_to = "Ethnicity", values_to = "Proportion") %>%
  head(140) # 20 schools with lowest admission rate

ggplot(prop_ad, aes(x = INSTNM, y = as.numeric(Proportion), fill = Ethnicity)) +
  geom_bar(position="stack", stat="identity") + 
  coord_flip()+
  ggtitle("Proportion of Ethnicity in Institution with Lowest Admission Rate") +
  xlab("Institution") +
  ylab("Ethnicity Proportion")
```

```{r number of schools of blue state, echo=FALSE}
blue = college %>% 
  filter(college$STABBR %in% c("WA", "OR", "CA", "HI", "IL", "MD", "DE", "NJ", "CT", "RI", "MA", "NY", "VT", "NH", "ME"))

num_school_blue = blue %>%
  group_by(STABBR) %>%
  summarize(count = n())

ggplot(data = num_school_blue,
       aes(x = STABBR, y = count, fill=STABBR)) +
  geom_col() +
  geom_text(aes(x=STABBR, y=count-5, label=count), size=3.5) + 
  ggtitle("Number of Schools in a more liberal states") +
  xlab("State")+
  ylab("Number of Institutions")
```


```{r number of schools of red state, echo=FALSE}
red = college %>% 
  filter(college$STABBR %in% c("AK", "MT", "ND", "ID", "WY", "SD", "UT", "MO", "KS", "OK", "LA", 'MS', 'AL', 'GA', 'TN', 'SC', 'WV', 'IN', 'NM', 'AZ', 'TX', 'KY', 'OH', 'NC'))

num_school_red = red %>%
  group_by(STABBR) %>%
  summarize(count = n())

ggplot(data = num_school_red,
       aes(x = STABBR, y = count, fill=STABBR)) +
  geom_col() +
  geom_text(aes(x=STABBR, y=count-5, label=count), size=3.5) + 
  ggtitle("Number of Schools in a more conservative states") +
  xlab("State")+
  ylab("Number of Institutions")
```

## Discussion

We would like to know whether there is a correlation between proportion of ethnicity and average sat score

```{r, echo=FALSE, warning = FALSE, message = FALSE}
score = college %>%
  group_by(INSTNM) %>%
  select("INSTNM", "SAT_AVG", "UGDS_WHITE", "UGDS_BLACK", "UGDS_ASIAN") %>%
  summarize(across(everything(), mean)) %>%
  mutate(other = 1 - UGDS_WHITE - UGDS_BLACK - UGDS_ASIAN)

## Multiple regression model
fit = lm(SAT_AVG ~ UGDS_WHITE + UGDS_BLACK + UGDS_ASIAN, data = score)

qqnorm(resid(fit), main = "Q-Q Plot of resid of multiple regression model")

## Predicted vs. observed avg sat score
score %>% 
  add_predictions(fit) %>% 
  ggplot(aes(x = SAT_AVG, y = pred)) +
  geom_point() +
  geom_smooth(method = "lm") +
  labs(x = "Average SAT score", y = "Predicted average SAT score", title = "Predicted vs. observed avg sat score")
## Residual plot
score %>% 
  add_residuals(fit) %>% 
  ggplot(aes(x = SAT_AVG, y = resid)) +
  geom_point() +
  geom_hline(yintercept = 0, color = "red", linetype = "dashed") + 
  labs(x = "Predicted SAT average score", y = "Residuals", title = "Residual of predicted avg sat score")


summary(fit)

```
The linear model fitted is given by y_hat = 990.03 + 182.04*white - 110.74*black + 1190.96*asian

Hypothesis testing:
$H_o =$ There is no effect on average SAT score due to proportion of ethnicity (beta_white = beta_black = beta_asian = 0)

$H_a =$ There is some effect on average SAT score due to proportion of ethnicity, at least one of the proportion have effect on avg SAT score (beta_white != 0 || beta_black != 0 || beta_asian != 0).

Since F = 302.6 where p < 2.2e-16, at the significance of alpha = 0.05 we can reject the null hypothesis that there is no effect on average SAT score due to proportion of ethnicity.


```{r, echo=FALSE, warning = FALSE, message = FALSE}
# plot
white = ggplot(score, aes(x = UGDS_WHITE, y = SAT_AVG)) +
  geom_point() +
  xlab("White Proportion") +
  ylab("SAT Average") +
  ggtitle("White vs SAT") +
  geom_smooth(se = FALSE, method = "lm") +
  theme_bw() 
qqnorm(resid(lm(SAT_AVG ~ UGDS_WHITE, data = score)), main = "White error Normal Q-Q Plot")

qqnorm(resid(lm(SAT_AVG ~ UGDS_WHITE, data = score)), main = "White error Normal Q-Q Plot")

black = ggplot(score, aes(x = UGDS_BLACK, y = SAT_AVG)) +
  geom_point() +
  xlab("Black Proportion") +
  ylab("SAT Average") +
  ggtitle("Black vs SAT ") +
  geom_smooth(se = FALSE, method = "lm") +
  theme_bw() 
qqnorm(resid(lm(SAT_AVG ~ UGDS_BLACK, data = score)), main = "Black error Normal Q-Q Plot")

qqnorm(resid(lm(SAT_AVG ~ UGDS_BLACK, data = score)), main = "Black error Normal Q-Q Plot")

asian = ggplot(score, aes(x = UGDS_ASIAN, y = SAT_AVG)) +
  geom_point() +
  xlab("Asian Proportion") +
  ylab("SAT Average") +
  ggtitle("Asian vs SAT ") +
  geom_smooth(se = FALSE, method = "lm") +
  theme_bw()
qqnorm(resid(lm(SAT_AVG ~ UGDS_ASIAN, data = score)), main = "Asian error Normal Q-Q Plot")

qqnorm(resid(lm(SAT_AVG ~ UGDS_ASIAN, data = score)), main = "Asian error Normal Q-Q Plot")

other = ggplot(score, aes(x = other, y = SAT_AVG)) +
  geom_point() +
  xlab("Other Ethnicities Proportion") +
  ylab("SAT Average") +
  ggtitle("Other Ethnicities vs SAT ") +
  geom_smooth(se = FALSE, method = "lm") +
  theme_bw() 

qqnorm(resid(lm(SAT_AVG ~ other, data = score)), main = "Other error Normal Q-Q Plot")

(white + black + asian) / other

# correlation coefficient
ws = score %>% pull(SAT_AVG)
prop_w = score %>% pull(UGDS_WHITE)
cat("white vs SAT correlation coefficient:", cor(ws,prop_w))

bs = score %>% pull(SAT_AVG)
prop_b = score %>% pull(UGDS_BLACK)
cat("black vs SAT correlation coefficient:", cor(bs,prop_b))

as = score %>% pull(SAT_AVG)
prop_a = score %>% pull(UGDS_ASIAN)
cat("asian vs SAT correlation coefficient:", cor(as,prop_a))

other = score %>% pull(SAT_AVG)
prop_o = score %>% pull(other)
cat("other ethnicity vs SAT correlation coefficient:", cor(other, prop_o))


```
The graph above show that white, asian and other ethnicity have positive correlation while black ethnicity has negative correlation.


**Shortcomings:** 

Since we don't have the admission score for each student but instead an average of all students, there could be bias if we use proportion of ethnicity vs SAT score. For example, one institution can have higher white proportion but the minority students who are not white might score very high so that the average admission score is high.

*Next, we will perform a hypothesis testing on the average score of sat between red states and blue states*

**Hypothesis testing:**

**Population and Sample:** 

Population -> Average SAT score of each institution in the US

Sample -> Average SAT score of each institution during 2017-2021 where admission score is not missing (3215 samples)

**Statistical Model:**

Institutions are $x_1,…,x_n$ for $n=3215$

Model these institutions as a random sample from the larger population

Let $F$ be this unspecified distribution

Let $μ$ be the mean

Let $σ$ be the standard deviation

$Xi∼F(μ,σ),i=1,…,n$

$Yi∼F2(μ2,σ2),i=1,…,n2$

**State Hypothesis:**

$H_o =$ There is no difference between average sat score of a more liberal states and a more conservative states

$H_a =$ There is a difference between average sat score of a more liberal states and a more conservative states

**Choose a Test Statistic:**

Test statistic $t=(\overline{x}−\overline{y})/(SE(\overline{x}−\overline{y}))$

```{r, echo=FALSE}
red_score = red %>%
  group_by(INSTNM) %>%
  summarize(sat = round(mean(SAT_AVG))) %>%
  pull(sat)
  

blue_score = blue %>%
  group_by(INSTNM) %>%
  summarize(sat = round(mean(SAT_AVG))) %>%
  pull(sat)

t.test(blue_score, red_score)
```

**Interpretation**

We are 95% confident that the average sat score of a more liberal state is between 43.98159 and 79.56565 higher than the average sat score of a more conservative state.

**Potential future directions for additional work**

- If we could find the admission score of each student and the institution they attend, we could better understand how each ethnicity performs

- After collecting individual student's data, we could see if blue states have higher diversity in ethnicity than red states

**Conclusion**